Thesis: draft

### 2021-05-08 draft.tex
latest version

\restoregeometry
\selectlanguage{english}
\newgeometry{top=20mm, bottom=25mm}
\tableofcontents
\addcontentsline{toc}{chapter}{Table of Contents}
\listoffigures
\addcontentsline{toc}{chapter}{List of Figures}
\listoftables
\addcontentsline{toc}{chapter}{List of Tables}
\restoregeometry

@@

The formation of echo chambers versus the fraction of completely stubborn actors, fracRes, in the Last.fm on-line social network is studied. The results are compared to corresponding results from network models, all with $N = 8003$: (i) the Watts-Strogatz model with $\left<k\right> = 4$, $\beta = 0.15$ and $\left<cc\right> = 0.317 \pm 0.004$; (ii) the stochastic block model with 53 communities of size 151, $p_\text{cl} = 0.023$, $p_\text{add} = 0.0001$, $\left<k\right> = 4.24 \pm 0.03$ and $Q = 0.811 \pm 0.004$; (iii) the stochastic Watts-Strogatz block model with 53 communities of size 151, each community has the following Watts-Strogatz parameters: $\left<k\right> = 4$, $\beta = 0.025$; edges are added between communities with probability $p_\text{add} = 0.0001$, $\left<k\right> = 4.77 \pm 0.01$, $\left<cc\right> = 0.344 \pm 0.002$ and $Q = 0.833 \pm 0.003$. The results for the PR and REC filtering algorithm are shown in Fig. \ref{echo_vs_fracRes_real}.\\
\newline
The PR filtering algorithm shows a decrease in the formation of echo chambers with an increasing fraction of stubborn actors. This is valid for both the real-world network and the network models. The stochastic Watts-Strogatz block model has a similar clustering and community structure as the Last.fm network. The Last.fm network has, however, a smaller formation of echo chambers than the SBM-WS. This suggests that it is not only clustering and/or community structure that drives the formation of echo chambers. Other topologies must play a role as well. The biggest difference between the real-world Last.fm network and the stochastic Watts-Strogatz block model lies in the degree distribution: the Last.fm network has a heterogeneous degree distribution instead of a homogeneous degree distribution. It is probably this heterogeneous contact pattern that hampers the formation of echo chambers in the real-world network compared to a network model with similar clustering and modularity, but with a homogeneous degree distribution. This agrees with the result that Perra and Rocha \cite{Perra2019} found: a heterogeneous contact pattern hampers the formation of echo chambers. % talk about possible effect of powerlaw community sizes + figure??
The results for the Watts-Strogatz model are similar to those for the stochastic Watts-Strogatz block model. The formation of echo chambers seems to be slightly smaller in the WS than in the SBM-WS network. It may be tempting to attribute this to the lack of community structure in the WS model. It must however be noted that the implemented WS model has a slightly smaller clustering coefficient ($\left<cc\right> = 0.317 \pm 0.004$) than the SBM-WS ($\left<cc\right> = 0.344 \pm 0.002$), which may also explain the somewhat smaller formation of echo chambers in the WS network than in the SBM-WS network. Furthermore, it is interesting to notice that the Last.fm on-line social network coincides with the SBM with a similar modularity. It seems that the hampering of the formation of echo chambers due to a heterogeneous contact pattern is of the same size as the hampering due to a lack of high clustering. % talk about coincidence??

\begin{figure}[H]
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{REC}
  	\includegraphics[width=8cm, height=7cm]{echo_chamber_vs_fracRes_PR_real_network_8x7.png}
    \label{pr_real}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{PR}
  	\includegraphics[width=8cm, height=7cm]{echo_chamber_vs_fracRes_REC_real_network_8x7.png}
    \label{rec_real}
  \end{subfigure}
  \captionsetup{format=plain}
  \caption[Formation of echo chambers versus fraction of completely stubborn actors for the PR and REC filtering algorithms and an initial $50/50$ opinion distribution. The Last.fm on-line social network is compared to corresponding network models.]{Formation of echo chambers versus the fraction of completely stubborn actors, fracRes, for an initial $50/50$ opinion distribution. The Last.fm on-line social network is compared to corresponding network models: (i) WS with $\left<k\right> = 4$, $\beta = 0.15$ and $\left<cc\right> = 0.317 \pm 0.004$; (ii) SBM-WS with 53 communities of size 151, each community has the following Watts-Strogatz parameters: $\left<k\right> = 4$, $\beta = 0.025$; edges are added between communities with probability $p_\text{add} = 0.0001$, $\left<k\right> = 4.77 \pm 0.01$, $\left<cc\right> = 0.344 \pm 0.002$ and $Q = 0.833 \pm 0.003$; (iii) SBM with 53 communities of size 151 with $p_\text{cl} = 0.023$, $p_\text{add} = 0.0001$, $\left<k\right> = 4.24 \pm 0.03$ and $Q = 0.811 \pm 0.004$. \textbf{(a)} PR filtering algorithm; \textbf{(b)} REC filtering algorithm. The $y$-axis shows the average of the normalized fraction of nodes with all neighbors having opinion A and the normalized fraction of nodes with all neighbors having opinion B. This value represents the formation of echo chambers in the system (denoted with echo chamber size in the figure). Each data point is the average of $50$ independent simulations. Bars represent the standard deviation $\sigma$.}
\label{echo_vs_fracRes_real}
\end{figure}
\noindent
The REC filtering algorithm enforces an increase in the formation of echo chambers when a small fraction of stubborn actors is introduced. If this fraction is further increased, the formation of echo chambers starts to decrease again. The increase is the strongest for the high clustering network models (WS, SBM-WS). Even though the Last.fm network has a similar clustering as these high clustering models, the formation of echo chambers and the corresponding increase in that formation are significantly smaller. As for the PR filtering algorithm, this is probably a consequence of the heterogeneous degree pattern of the Last.fm network. % talk about SBM

@@

% talk about evolution of opinions??

@@

Finally, it was found that the initial distribution of the opinions over the communities in a network with community structure (SBM) does make a difference. We investigated a scenario in which a fraction of the communities was given a single opinion, opinion A, while the remaining communities were given an initial $50/50$ opinion distribution. This scenario was then compared to the case in which the total opinion distribution was randomly distributed over the system. This was done for both a high modularity and a low modularity network and for both the REC and PR filtering algorithms. \\

@@

For the PR filtering algorithm, the randomly distributed scenario resulted in an increase in the formation of echo chambers around the majority opinion A. This was observed for both the low and high modularity networks. The community distributed scenario for the low modularity network hampers this increase. It was found that this is more than just a result of the increased number of initial opinion A echo chambers. For the high modularity network, the community distributed scenario showed an even larger hampering of the formation of echo chambers around the majority opinion A. This is probably the result of: (i) an even higher initial number of echo chambers around the majority opinion A than for the low modularity network; (ii) the hampering in the increase in the majority opinion A for the community distributed scenario in case of the high modularity network combined with the PR filtering algorithm. \\

@@

two dominant and opposing echo chambers'' (Wang et al., p.10, \cite{Wang2020}). This agrees with our result that the PR filtering algorithm, combined with stubborn actors, hinders the formation of echo chambers.

>>>>

\restoregeometry
\selectlanguage{english}
\tableofcontents
\addcontentsline{toc}{chapter}{Table of Contents}
\newgeometry{top=20mm, bottom=25mm}
\listoffigures
\addcontentsline{toc}{chapter}{List of Figures}
\listoftables
\addcontentsline{toc}{chapter}{List of Tables}
\restoregeometry

@@

The formation of echo chambers versus the fraction of completely stubborn actors, fracRes, in the Last.fm on-line social network is studied. The results are compared to corresponding results from network models, all with $N = 8003$: (i) the Watts-Strogatz model with $\left<k\right> = 4$, $\beta = 0.15$ and $\left<cc\right> = 0.317 \pm 0.004$; (ii) the stochastic block model with 53 communities of size 151, $p_\text{cl} = 0.023$, $p_\text{add} = 0.0001$, $\left<k\right> = 4.24 \pm 0.03$ and $Q = 0.811 \pm 0.004$; (iii) the stochastic Watts-Strogatz block model with 53 communities of size 151, each community has the following Watts-Strogatz parameters: $\left<k\right> = 4$, $\beta = 0.025$; edges are added between communities with probability $p_\text{add} = 0.0001$, $\left<k\right> = 4.77 \pm 0.01$, $\left<cc\right> = 0.344 \pm 0.002$ and $Q = 0.833 \pm 0.003$. The results for the PR and REC filtering algorithm are shown in Fig. \ref{echo_vs_fracRes_real}.\\
The PR filtering algorithm shows a decrease in the formation of echo chambers with an increasing fraction of stubborn actors. This is valid for both the real-world network and the network models. The stochastic Watts-Strogatz block model has a similar clustering and community structure as the Last.fm network. The Last.fm network has, however, a smaller formation of echo chambers than the SBM-WS. This suggests that it is not only clustering and/or community structure that drives the formation of echo chambers. Other topologies must play a role as well. The biggest difference between the real-world Last.fm network and the stochastic Watts-Strogatz block model lies in the degree distribution: the Last.fm network has a heterogeneous degree distribution instead of a homogeneous degree distribution. It is probably this heterogeneous contact pattern that hampers the formation of echo chambers in the real-world network compared to a network model with similar clustering and modularity, but with a homogeneous degree distribution. This agrees with the result that Perra and Rocha \cite{Perra2019} found: a heterogeneous contact pattern hampers the formation of echo chambers. Another difference between the Last.fm and the SBM and SBM-WS network is the community structure. The Last.fm network has a community structure where the community sizes follow a heterogeneous distribution: plenty of small communities and a few large ones. The stochastic block model and stochastic Watts-Strogatz block model, that we used, only have communities of a single size. This does, however, not play a significant role in the formation of echo chambers, as can be seen in Fig. \ref{power_vs_single}. In this figure, the distribution of echo chambers in the Last.fm network is compared to those in the SBM-WS network with similar clustering and modularity. The SBM-WS is modeled in two ways: (i) the communities all have a fixed size; (ii) the communities have the same heterogeneous sizes as in the Last.fm network. The PR and REC filtering algorithms are compared.\\ % add parameters of SBM-WS power??
The results for the Watts-Strogatz model are similar to those for the stochastic Watts-Strogatz block model. The formation of echo chambers seems to be slightly smaller in the WS than in the SBM-WS network. It may be tempting to attribute this to the lack of community structure in the WS model. It must however be noted that the implemented WS model has a slightly smaller clustering coefficient ($\left<cc\right> = 0.317 \pm 0.004$) than the SBM-WS ($\left<cc\right> = 0.344 \pm 0.002$), which may also explain the somewhat smaller formation of echo chambers in the WS network than in the SBM-WS network. Furthermore, it is interesting to notice that the Last.fm on-line social network coincides with the SBM with a similar modularity. It seems that the hampering of the formation of echo chambers due to a heterogeneous contact pattern is of the same size as the hampering due to a lack of high clustering. % talk about coincidence??

\begin{figure}[H]
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{PR}
  	\includegraphics[width=8cm, height=7cm]{echo_chamber_vs_fracRes_PR_real_network_8x7.png}
    \label{pr_real}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{REC}
  	\includegraphics[width=8cm, height=7cm]{echo_chamber_vs_fracRes_REC_real_network_8x7.png}
    \label{rec_real}
  \end{subfigure}
  \captionsetup{format=plain}
  \caption[Formation of echo chambers versus fraction of completely stubborn actors for the PR and REC filtering algorithms and an initial $50/50$ opinion distribution. The Last.fm on-line social network is compared to corresponding network models.]{Formation of echo chambers versus the fraction of completely stubborn actors, fracRes, for an initial $50/50$ opinion distribution. The Last.fm on-line social network is compared to corresponding network models: (i) WS with $\left<k\right> = 4$, $\beta = 0.15$ and $\left<cc\right> = 0.317 \pm 0.004$; (ii) SBM-WS with 53 communities of size 151, each community has the following Watts-Strogatz parameters: $\left<k\right> = 4$, $\beta = 0.025$; edges are added between communities with probability $p_\text{add} = 0.0001$, $\left<k\right> = 4.77 \pm 0.01$, $\left<cc\right> = 0.344 \pm 0.002$ and $Q = 0.833 \pm 0.003$; (iii) SBM with 53 communities of size 151 with $p_\text{cl} = 0.023$, $p_\text{add} = 0.0001$, $\left<k\right> = 4.24 \pm 0.03$ and $Q = 0.811 \pm 0.004$. \textbf{(a)} PR filtering algorithm; \textbf{(b)} REC filtering algorithm. The $y$-axis shows the average of the normalized fraction of nodes with all neighbors having opinion A and the normalized fraction of nodes with all neighbors having opinion B. This value represents the formation of echo chambers in the system (denoted with echo chamber size in the figure). Each data point is the average of $50$ independent simulations. Bars represent the standard deviation $\sigma$.}
\label{echo_vs_fracRes_real}
\end{figure}
\noindent
The REC filtering algorithm enforces an increase in the formation of echo chambers when a small fraction of stubborn actors is introduced. If this fraction is further increased, the formation of echo chambers starts to decrease again. The increase is the strongest for the high clustering network models (WS, SBM-WS). Even though the Last.fm network has a similar clustering as these high clustering models, the formation of echo chambers and the corresponding increase in that formation are significantly smaller. As for the PR filtering algorithm, this is probably a consequence of the heterogeneous degree pattern of the Last.fm network. The results for the Last.fm network coincide with those of the SBM. As for the PR filtering algorithm, this seems to suggest that the hampering in the formation of echo chambers due to a heterogeneous degree distribution is of a similar size as the corresponding hampering due to the lack of high clustering.

\begin{figure}[H]
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{REC}
  	\includegraphics[width=8cm, height=7cm]{echo_chambers_SBM_powerlaw_vs_regular_REC.png}
    \label{pr_pow_single}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{PR}
  	\includegraphics[width=8cm, height=7cm]{echo_chambers_SBM_powerlaw_vs_regular_PR.png}
    \label{rec_pow_single}
  \end{subfigure}
  \captionsetup{format=plain}
  \caption[Distribution of the fraction of friends (nearest neighbors, nn) of node $i$ with the same opinion B at $t = 500$. The SBM-WS with heterogeneous community sizes is compared to the SBM-WS with a single community size. Results are compared to the real-world Last.fm network with similar clustering and modularity. Results for the REC and PR filtering algorithms.]{Distribution of the fraction of friends (nearest neighbors, nn) of node $i$ with the same opinion B at $t = 500$. The distribution is normalized by dividing for the same quantity computed at $t=0$. The SBM-WS with heterogeneous community sizes is compared to the SBM-WS with a single community size. Results are compared to the real-world Last.fm network with similar clustering and modularity. \textbf{(a)} REC filtering algorithm; \textbf{(b)} PR filtering algorithm. Each curve is the average of $50$ independent simulations. Error bars are omitted for an improved visualization.}
\label{power_vs_single}
\end{figure}

@@

@@

These results give us an insight in the interplay between network structure, filtering algorithms and individual resistance. These insights may be crucial to understand phenomena such as polarization in society and to determine the impact of social media on these phenomena. They learn us that even though semantic filtering (PR) strongly enhances the formation of echo chambers in the absence of stubborn actors, this enhancing does not stand on its own. An interplay between a network with a high clustering and/or community structure is needed to effectively see this increase in echo chambers induced by semantic filtering. A random network, combined with semantic filtering, does not enhance the formation of echo chambers. Furthermore, if there are no stubborn actors, recent or random filtering (REC/REF) seems to be more protective against the formation of echo chambers than semantic filtering. If there are, however, small fractions of stubborn actors in the system, this scenario changes: semantic filtering starts to hamper the formation of echo chambers, whereas recent or random filtering begins to enhance polarization. It is often assumed that algorithmic personalization, tailored to ones preferences, is one of the main reasons of the appearance of echo chambers on social media \cite{Ge2020}\cite{Mohseni2018}\cite{Stark2020}. Our results may, however, suggest, that if there is some kind of stubbornness (which is probably the case in the real world), heterogeneous exposure (REF/REC) to different opinions may be more harmful than a homogeneous exposure (PR). Bail et al. found experimental results that may agree with this finding \cite{Bail2018}. This is an interesting result, that should be investigated more thoroughly.\\
Finally, it was found that the initial distribution of the opinions over the communities in a network with community structure (SBM) makes a difference. We investigated a scenario in which a fraction of the communities was given a single opinion, opinion A, while the remaining communities were given an initial $50/50$ opinion distribution. This scenario was then compared to the case in which the total opinion distribution was randomly distributed over the system. This was done for both a high modularity and a low modularity network and for both the REC and PR filtering algorithms. \\

@@

For the PR filtering algorithm, the randomly distributed scenario resulted in an increase in the formation of echo chambers around the majority opinion A. This was observed for both the low and high modularity networks. The community distributed scenario for the low modularity network hampers this increase. It was found that this is more than just a result of the increased number of initial opinion A echo chambers. For the high modularity network, the community distributed scenario showed an even larger hampering of the formation of echo chambers around the majority opinion A. This is probably the result of: (i) an even higher initial number of echo chambers around the majority opinion A than for the low modularity network; (ii) the hampering in the increase of the majority opinion A for the community distributed scenario in case of the high modularity network combined with the PR filtering algorithm. \\

@@

two dominant and opposing echo chambers'' (Wang et al., p.10, \cite{Wang2020}). This agrees with our result that the PR filtering algorithm, combined with stubborn actors, hinders the formation of echo chambers.\\
It must be emphasized that there are limitations to our model and methods. As already stated before, the simpleness of the model, which on one hand can be seen as a strength, may also be one of the weaknesses. Real-life social behavior and opinion dynamics on social media may be far from simple. As a consequence, there may be many dynamics of real-life opinion formation that are not well-captured by our model. More empirical data is therefore needed to compare the results, obtained from our model, against. Other limitations may be the rather small system sizes that were used in our simulations ($N = 10^3$). It is unclear how the system size affects the results and it is recommended that the model is tested on both larger network models and real-world networks. Another possible shortcoming is the binary nature of the model. There are real-life cases in which people have to choose between two competing opinions. These situations, however, do not cover the complete range of possible opinion dynamics situations. It would be interesting to run the model on situations where more than two opinions are allowed or to switch to continuous opinion dynamics. We also did not differentiate between the agents in our model, except for non-stubborns versus stubborns. This combined with the homogeneous degree distribution of our networks models may be another limitation. Real-life social media platforms display a heterogeneous contact pattern, where the `hubs' are often more influential than other users, not only by their larger reach, but also by their status. We could adapt our model to mimic this real social media behavior more closely. Therefore, a heterogeneous network model should be introduced, combined with an influence parameter. This parameter should be used to determine the influence of an agent on others: the opinion of a more influential agent is easier adapted by the others. In the model, the high degree nodes should then be made more influential. As a final limitation, we mention the fixed stubbornness of the agents. Even though the stubbornness parameter $r$ could be individualized to each node, we did not include the possibility to be more stubborn towards opinions coming from one friend than from another. This idea may be linked to the influence of certain friends/acquaintances: somebody may adapt another viewpoint more easily from certain friends than from others.\\
Some future investigation paths, linked to limitations of our model, were given in the previous paragraph. Other future directions may be: (i) investigating the distribution of the stubborn actors (community distributed versus random); (ii) making one opinion more stubborn than the other; (iii) investigating the effect of the activation mechanism: what happens if we switch to the heterogeneous activation mechanism (see Section \ref{actMech})? and (iv) making the network temporal in the sense that it becomes possible to add and remove friends. Sasahara et al. investigated a model in which social influence was combined with social rewiring (unfriending) \cite{Sasahara2020}. Their findings were that these processes of influence and unfriending enhanced the formation of echo chambers \cite{Sasahara2020}. It might be interesting to incorporate their model features into our model with algorithmic personalization and stubbornness.
